---
subtitle: "Stats 306: Lecture 19"
title: "Iteration Continued"
output: 
  learnr::tutorial:
    progressive: true
    css: css/lecture.css
runtime: shiny_prerendered
---


```{r setup, include=FALSE}
library(learnr)
library(tidyverse)


wikiart <- read_tsv("./data/WikiArt-Emotions-All.tsv.gz")
wikiart <- mutate(wikiart, Year = as.numeric(str_sub(Year, 1, 4)))

pivot_longer(wikiart, 
             matches(":"), 
             names_to = c("rating_type", "emotion"),
             names_sep = ":",
             values_to = "mean_annotation") -> wa_long

select(wa_long, Title, rating_type, emotion, mean_annotation)
wa_long <- mutate(wa_long,
                  emotion = str_replace(emotion, "\\s", ""))

filter(wa_long, str_detect(rating_type, "Art")) |>
  group_by(ID) |>
  arrange(desc(mean_annotation)) |>
  summarize(strongest_emotion = first(emotion),
            strongest_emotion_value = first(mean_annotation)) ->
  wa_art_strongest

wa_art_strong_year <- left_join(wikiart, wa_art_strongest, by = "ID")
```

## Review

* Vectorization: performing same operation on all items in a collection
* Iteration: one at a time computations over a collection
* `for` loops
  * `for (item in collection) { ... }`
  * often have to set up storage in advance
  * very flexible
* `while` loops until condition met
* `map`: perform and save one value per input collection
* `map` (list) vs. `map_*` (vectors of given types)
* `map2` and `map2_*`
* Use of formulas as functions: `~ ... do things with .x ...`

## Compound Random Variables

Many times we take sample sizes $n$ to be fixed. But what if our sample size itself was random? 

A **compound random variable** is one in which a discrete random variable sets the sample size, and then some other random variable is replicated that many times.

```{r}
r_pois_norm <- function(n, lambda = 1, mu = 0 , sd = 1) {
  replicate(n, {
    n_i <- rpois(1, lambda) + 1
    rnorm(n_i, mu, sd)
  })
}

samples <- r_pois_norm(3, lambda = 8)
samples
```


## Mapping on two things

Suppose for each sample, we have another outcome:

```{r}
samples_y <- map(samples, ~ runif(length(.x)))
samples_y
```

What is the difference correlation within each sample?

```{r}
map2(samples, samples_y, cor)
map2_dbl(samples, samples_y, cor)
```


## Mapping your own function on two things

Suppose we wanted the average ratio of X to Y within each sample. We need to write a custom function with two arguments:

```{r}
map2_dbl(samples, samples_y, function(x, y) { mean(x / y)})
map2_dbl(samples, samples_y, ~ mean(.x / .y))
```

## Exercise

Using `map2`, get the first word of each item in `stringr::fruit` (keep single words). 

Some useful examples:
```{r}
str_locate("chili pepper", " ")
str_sub("chili pepper", end = 6 - 1)
str_locate("canataloupe", " ")
```


```{r firstword, exercise = TRUE}
first_space <- str_locate(stringr::fruit, " ")[,1]
```


## Other loop replacing functions

`keep` takes only the entries that match a predicate:

```{r}
keep(samples, ~ length(.x) > 10)
```

`discard` drops entries that match.

Many `for` loops have **accumulators**:

```{r}
max_length <- -Inf
for (i in seq_along(samples)) {
  max_length <- max(length(samples[[i]]), max_length)
}
max_length
```

`reduce` avoids the bookkeeping (often convenient to map first):
```{r}
map_dbl(samples, length) |> reduce(max)
```

We can see the intermediate results with `accumulate`:
```{r}
accumulate(c(1, 10, 100, 2, 3, 4), `+`)
```

## Map functions that return tables

The `_dfr` variants use functions that return data frames, and then row bind them together
```{r}
map_dfr(samples, ~ tibble(x = .x))
```
We probably want to know which sample each value is so we can do operations after. We can use an **indexed map** with the `imap_*` variants:
```{r}
imap_dfr(samples, ~ tibble(x = .x, sample_num = .y)) |> 
  group_by(sample_num) |> 
  mutate(x_centered = x - mean(x)) # within group, mean centered values
```

There is also `_dfc` that binds columns together.


## Connections to grouping and summary functions

Many of these tasks can also be done by creating tables and running `group_by` and `summarize`. There is also a `group_map`

```{r}
group_by(wikiart, Category) |> group_map(function(x, grp) { 
    if (nrow(x) > 10) {
      return(mean(x$`Ave. art rating`))
    } else {
      return(max(x$`Ave. art rating`))
    }
  }) |> flatten_dbl()
```

## Built in R functions

There are several built in functions that are similar to `map_*`:

* `lapply` is similar to `map`, but doesn't allow for the function shorthand notation
* `sapply` is similar to the `map_*` varieties, but no checking. Will return a list if it can't safely unlist results.
* `vapply` is even closer to `map_*`, but still no anonymous functions
* `mapply` is similar to `map2`

R also has some capitalized versions (`Map`, `Reduce`), but generally no checking less convenient to use with pipes.

## Errors in iteration

If we run into a bug in the code being iterated, it can often be difficult to find the source.
```{r}
x <- list(1, 10, "3", 7)
```
```{r eval = FALSE}
map(x, log)
# Error in .Primitive("log")(x, base) : 
#  non-numeric argument to mathematical function
```

The `safely` function catches these errors and gives back useful info:

```{r}
safe_log <- safely(log) # returns a new function
safe_log(10)
safe_log("3")
map(x, safe_log)
```


## `transpose` function

Recall our model of a table:  a list of length n composed of vectors of length k. If we wanted to **transpose** the table we would have a list of length k with vectors of length n.

The `transpose` function does this for data that can be thought of as a table:
```{r}
x_log_tp <- transpose(map(x, safe_log))
x_log_tp$result
x_log_tp$result |> flatten_dbl()
x_log_tp$error
```

