---
subtitle: "Stats 306: Lecture 24"
title: "Shiny Server and Reactive Programming"
output: 
  learnr::tutorial:
    progressive: true
    css: css/lecture.css
runtime: shiny_prerendered
---


```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(lubridate)
library(shiny)
```

## Review

* Shiny: Dynamic display of data
* Three parts: UI, Server, Reactive Programming
* UI
  * consistent layout of app (usually composed using `fluidPage` and `fluidRow`)
  * place `input*` and `output*` with ids that map to the `input$ID` and `output$ID` in `server`
  * interactive graphics
  * basic text and number input
  * sliders for ranged values
* Widget gallery is a nice view of the different basic tools
* [Shiny Homepage](https://shiny.rstudio.com/) (the gallery is a neat set of examples)
* [Mastering Shiny](https://mastering-shiny.org/index.html)

## Inputs for categorical data

When we have a set of categories we want for input, we have two options:

* Radio buttons: only one can be selected
* Check boxes: many can be selected

In both cases, we'll give a set of choices of the form: `c("Nice display name 1" = "option1", "Nice display name 2" = "option2", ...)`

See `checkbox_mpg.R`.

## Exercise

Use `radioButtons` (same basic interface as `inputCheckboxGroup`) to select two numeric variables from the `mpg` table to graph against each other.

Bonus: if the selected variables are the same, graph as a density instead of a jitter.

See `example_radio.R`.

## `outputXYZ` and `renderXYZ` functions

Once we get the input we need, it's time to get it back to the user. Within the UI, we place the output using various `outputXYZ` functions:

* `htmlOutput()/uiOutput()`
* `imageOutput()` (static image from a file)
* `plotOutput()` (R plots)
* `textOutput()` HTML text
* `verbatimTextOutput()` non-rendered text
* `tableOuput()/dataTableOutuput()` tables (`dataTable` version has interactive elements)
* several others for download links, dialog boxes, etc

We need to create these outputs using corresponding `renderXYZ` functions

* `renderUI`
* `renderImage`
* `renderPlot`
* `renderPrint/renderText` (print verison captures output, text version takes a string vector as the return value)
* `renderTable/renderDataTable`

## Including HTML in output

If you know some basic HTML, you can include it the output.

The `ul` tag starts and ends bulleted lists. `li` tags delimit individual items in the list.

See `input_song_html.R`

## Including Markdown

See `input_song_markdown.R`

## Shiny in RMarkdown documents

Go to RStudio...

## Exercise

Open `exercise_markdown.R` and follow the comments.

## Manipulating user inputs

When users update values on inputs this generates **events**. We get notifications of events using `observeEvent(input$element, { code })`. Then we can use `update*` functions.

See `slider_update_update.R`

## Events exercise

Open `exercise_dynamic_radios.R`

## Making web apps that feel like real apps

Web App Goal: make a series of web pages feel like an interactive application

Why is this difficult?

* Client/server architecture
* Keeping state persistent
* Avoid having to transfer lots of data
* Avoid re-running lots of computations on server
* Keep manageable for the programmer

## Reactive programming

We already saw **event driven programming** where we could respond to changes to UI.

In **reactive programming** we also allow variables to react to changes in other variables.

Downside: we have to use a few more tools than just regular R variables.

## Reactive variables: function like objects

Variables (boxes to hold data) in Shiny's reactive framework look and behave much like functions.

```{r}
reactiveConsole(TRUE) # use this for interactive testing

rv <- reactiveVal(999) # initial value
rv() # get the value
rv(1) # set the value
rv()
```

## Reactive contexts

You may recall that when we use `render*` functions or `update*` functions we executed code inside of `{...}`. This formed a **reactive context**. A wrapper around code that connects reactive variables.

```{r}
rv2 <- reactive({
  message("updating rv2")
  rv() - 10
})

# get inital value
rv2()

# now value is stored
rv2() + 1

# does not immediately update rv2
rv(123)

# only runs code when needed
rv2()
```

## Example: Two sample problems (courtesy "Mastering Shiny")

In statistics, "two sample problems" are when we have samples from two populations and want to test if the populations are the same. Here is a function to plot the two samples and a function to test (using a t-test) if the populations have the same mean.

```{r}
freqpoly <- function(x1, x2, binwidth = 0.1, xlim = c(-3, 3)) {
  df <- data.frame(
    x = c(x1, x2),
    g = c(rep("x1", length(x1)), rep("x2", length(x2)))
  )

  ggplot(df, aes(x, colour = g)) +
    geom_freqpoly(binwidth = binwidth, linewidth = 1) +
    coord_cartesian(xlim = xlim)
}

t_test <- function(x1, x2) {
  test <- t.test(x1, x2)

  sprintf(
    "p value: %0.3f\n[%0.2f, %0.2f]",
    test$p.value, test$conf.int[1], test$conf.int[2]
  )
}
```

In action:

```{r}
x1 <- rnorm(100, mean = 0, sd = 0.5)
x2 <- rnorm(200, mean = 0.15, sd = 0.9)

freqpoly(x1, x2)
cat(t_test(x1, x2))
```

## Naive attempt to use in Shiny app

See `two_sample_01.R`.

## Saving results in reactive variable

We want both the graph and the t-test to use the same values, so we need for both outputs to have access to the same data. We can do that by populating a **reactive variable**.

Also, it would be nice if the data did not change if we only manipulate the characteristics.

See `two_sample_02.R`

## Additional reactive dependencies

It might be nice to refresh the graph with a button press. We can do this by adding a button to the UI and adding a dependency in the `reactive()` function call on the `input$button` value.

See `two_sample_03.R`

## Exercise

Open `exercise_reactive.R`

## `reactiveEvent`

In order to get access to user actions, we can use the `reactiveEvent` function.

## Shiny Next Steps

Lots more to learn!

- More reactive programming tools
- Tabs/multipage apps
- Add on packages
- Deployment to the web

