---
subtitle: "Stats 306: Lecture 15"
title: "Relational Data, Continued; Introduction to Strings"
output: 
  learnr::tutorial:
    progressive: true
    css: css/lecture.css
runtime: shiny_prerendered
---


```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
set.seed(2939394)
if (!require(nycflights13)) install.packages("nyclfights13", repos = "http://cran.us.r-project.org")

sonnet18 <- "Shall I compare thee to a summer's day?\nThou art more lovely and more temperate:\nRough winds do shake the darling buds of May,\nAnd summer's lease hath all too short a date;\nSometime too hot the eye of heaven shines,\nAnd often is his gold complexion dimm'd;\nAnd every fair from fair sometime declines,\nBy chance or nature's changing course untrimm'd;\nBut thy eternal summer shall not fade,\nNor lose possession of that fair thou ow'st;\nNor shall death brag thou wander'st in his shade,\nWhen in eternal lines to time thou grow'st:\n\tSo long as men can breathe or eyes can see,\n\tSo long lives this, and this gives life to thee."

sonnet18_lines <- str_split(sonnet18, "\n")[[1]]
```

## Review

* `nycflights` data: flights, airlines, airports, planes, weather
* Keys: **primary keys** identify data within the table, **foreign keys** link to data in other tables.
* Relations: tables created by **joining** two or more tables. Classify as: mutating or filtering.
* Types of joins: outer: left, right, and full; inner
* Use of `dplyr` `*_join` functions, `by` argument, `suffix` argument
* Compound keys, natural joins (no `by` argument)

## Other set operations functions

Base R has a series of **set operation** functions that operate on collects as mathematical sets (collections where items can appear at most once): `intersect`, `union`, `setdiff`, `setequal`, `is.element`, `%in%`

I frequently use these for doing comparisons of vectors:
```{r}
"DTW" %in% top_dest$dest # same as is.element
intersect(c("BOS", "LAX", "DTW"), top_dest$dest)
union(c("A", "B", "C"), c("Q", "A", "R"))
setdiff(c("A", "B", "C"), c("Q", "A", "R"))
setequal(c("A", "B", "C"), c("B", "C", "A"))
```

## Set operations with tables

With filtering joins, we can choose which columns link the tables. With the set operations, we must use all columns.

Flights to top destinations that are also delayed:
```{r}
flights_to_top <- semi_join(flights, top_dest) 
flights_delayed <- filter(flights, dep_delay > 0)
intersect(flights_to_top, flights_delayed)
```

## Relational data wrap up

* When data are stored in more than one table you need to **join**
* **Mutating joins** join tables to add columns (and sometimes rows!)
* Outer left/right/full joins, inner joins
* Also useful for linking table to itself
* Set operations less generally useful, but have some specific uses (e.g., vectors)

## Strings and Factors

R has two data types for representing categorical/nominal data types:

* Strings: character data
* Factors: integers representing categories with look up table for labels, can be ordered

```{r}
example_str <- c("you", "say", "goodbye", "i", "say", "hello", "hello", "hello")
example_str
example_fct <- factor(example_str)
example_fct
```

## Strings vs. factors

Many times two types have the same function, why use one or the other?

- Want to order lexicographically: strings
- Want to order in other way: factors
- Treat as text (search, combine): strings
- Avoid errors from treating as text: factors

There used to be a space saving advantage to factors but that has been eliminated with newer versions of R.

## String basics

A string (or "character vector" in R) is character data enclosed in double `"` or single `'` quotes.

To include quotation marks inside, we need to **escape** them (like `,` in CSVs):
```{r}
baberuth <- "George Herman \"Babe\" Ruth Jr., a.k.a. \"the Bambino\" and \"the Sultan of Swat\""
cat(baberuth) # prints directly instead of displaying as escaped string
```

## Other special characters

Since `\` is used to escape things, if you need a literal backslash use `\\`.
* New line: `\n`
* Tab: `\t`
* Unicode characters: `\unnnn` (u followed by 4 digit code)

This often come up in searching text for certain patterns.

## Combining strings

If we need to make strings out of other strings, we have two (more or less identical) functions:

* Old school: `paste` (joins with " ") and `paste0` (joins with ""); I frequently use these
* New school: `str_c` (joins with ""); used in our book

```{r}
paste("Section", c("100", "200", "300"),
      "meets on",
      c("Monday", "Tuesday", "Wednesday"))
```
Notice: this is a vector of length 3, use `collapse = "SEP"` to make a single string from a vector.

## Exercise

Join the following using a new line characters `\n` (use the `sep` and `collapse` arguments as necessary) and indent the last two lines using tabs (`\t`). Demonstrate using `cat`.

```{r sonnet18, exercise = TRUE}
line1 <- "Nor shall death brag thou wander'st in his shade,"
line2 <- "When in eternal lines to time thou grow'st:"
line3 <- "So long as men can breathe or eyes can see,"
line4 <- "So long lives this, and this gives life to thee."
```

## Properties of strings

```{r, echo = FALSE}
line1 <- "Nor shall death brag thou wander'st in his shade,"
line2 <- "When in eternal lines to time thou grow'st:"
line3 <- "So long as men can breathe or eyes can see,"
line4 <- "So long lives this, and this gives life to thee."
```

The `length` function in R will tell you how many items are in a vector:

```{r}
sonnet18_end <- c(line1, line2, line3, line4)
length(sonnet18_end)
```

To find the length of the strings in terms of characters:
```{r}
str_length(sonnet18_end) # base::nchar also works
```

## Subsetting

The `str_sub` function can retrieve or change **substrings**.

```{r}
str_sub(line1, 11, 20)
str_sub(line2, 36)
str_sub(line2, -8)
line4b <- line4
str_sub(line4b, 1, 10) <- "REMOVED"
line4b
```


## Splitting

Here's the full sonnet as a single string:
```{r}
sonnet18 <- "Shall I compare thee to a summer's day?\nThou art more lovely and more temperate:\nRough winds do shake the darling buds of May,\nAnd summer's lease hath all too short a date;\nSometime too hot the eye of heaven shines,\nAnd often is his gold complexion dimm'd;\nAnd every fair from fair sometime declines,\nBy chance or nature's changing course untrimm'd;\nBut thy eternal summer shall not fade,\nNor lose possession of that fair thou ow'st;\nNor shall death brag thou wander'st in his shade,\nWhen in eternal lines to time thou grow'st:\n\tSo long as men can breathe or eyes can see,\n\tSo long lives this, and this gives life to thee."
```

Let's split that up into individual lines as strings in a vector:
```{r}
sonnet18_lines <- str_split(sonnet18, "\n") # notice use of new line
sonnet18_lines
```

What is happening? Since `sonnet18` could have had multiple entries (vector), `str_split` returns a **list** with splits for each item in the vector:
```{r}
str_split(c("a b c", "d e"), " ")
```

In this case:
```{r}
sonnet18_lines <- sonnet18_lines[[1]] # [[]] for lists
```

## Exercise

**Tokenization** is the process many computer languages use to turn strings into discrete words. Use `str_split` to turn each line of `sonnet18_lines` into individual words. Use `unlist` on the result and answer the question: how many words are there?

```{r tokenize, exercise = TRUE}

```

