---
subtitle: "Data Science 306: Lecture 26"
title: "Packages and Object Oriented Programming"
output: 
  learnr::tutorial:
    progressive: true
    css: css/lecture.css
runtime: shiny_prerendered
---


```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
```

## Overview

In this lecture we will look at two topics:

* Packages: add-on functionality for R
* Object Oriented Programming: reuse of functions and data structures by defining classes of objects

## Packages

An **R package** is a collection of code and data that can be loaded using the `library` function. Library can use either "bare words" or quoted names.

```{r}
library(parallel)
library("boot")
```

You may also see `require` which allows you to handle if the package is not available in a graceful way:

```{r}

if (!require("package_that_does_not_exist")) {
  message("oops! not available!")
}
```
## Installing Packages

The Comprehensive R Archive Network provides hosting for packages. Installing is as simple as:

```{r, eval = FALSE}
install.packages("package_name") # notice use of quotes
```

You may have to select a host. Usually the "cloud" option is best.

Packages can also be installed from files on the drive, github, and arbitrary URLs. A useful helper to make this easier is the `devtools` package:

```{r, eval = FALSE}
install.packages("devtools")

library(devtools)
install_github("USERNAME/repo")
```


## Source and Binary packages

Many R packages have C/C++ or FORTRAN (an older scientific programming language) that needs to be **compiled** to programs that can run in your particular operating system.

For most operating systems, CRAN provides pre-compiled **binary** packages where the compilation has been done in advance. These are nice because you don't need the right compiler and set of C/C++ or FORTRAN libraries.

For other operating systems, or if you want to compile yourself, you can get **source** packages by adding the `type = "source"` option to `install.packages()`. Installing packages with C/C++ or FORTRAN from github will require the appropriate compilers.

Getting compilers set up is outside our scope, but [details can be found on CRAN](https://cran.r-project.org/).

## Package Dependencies

Packages can require or extend the functionality of other packages.

* Imports: these packages will always be installed along with this package. Critical functionality.
* Suggests: provides additional useful functionality, but the package will use `require` to make sure the package is available.
* Depends: Like imports but will load this package into the user's environment to make it available as well.

Usually, all will be installed as well when using `install.packages` but users can elect to skip Suggests.

## Visiting the Comprehensive R Archive Network (CRAN)

* [CRAN homepage](https://cran.r-project.org/)
* Example package: [Shiny](https://cran.r-project.org/web/packages/shiny/index.html)

## Creating your own packages

* Create a new Posit Cloud project
* Install `devtools` package
* Use the `devtools::create("name_of_package")
* More can be found at [R Packages by Wickham and Bryan](https://r-pkgs.org/)

## Object Oriented Programming

Everything we create in R is an "object", something that can be put into a variable/passed as an argument/returned from a function.

*Object oriented programming* explains how define types of objects and what we can do. There are many kinds of object oriented programming, but they commonly share:

* **Polymorphism**: "many shapes", many different types of objects can all be used in the same way
* **Inheritance**: we can define new types of objects by referencing existing types of objects.

We often call these "types" **classes**.

## Classes, fields, methods

* A **class** in R defines a structure for data. The components are called **fields**.
* R actually has several different ways of defining classes called S3, S4, and R6. The "S" indicates the predecessor to R and the number the sequential version number.
* A **method** is a version of a function that works for a particular class.

## Generic functions and methods

Some languages package methods within the classes. In R, we define **generic functions** that specify the interface for all classes, and then **methods** are versions of the function that work for specific data.

Example: `mean`:

```{r}
# generic
mean

# what methods exist?
methods(mean)

nums <- c(1, 15, -2, 6)
dates <- as.Date(c("2024-04-16", "1776-07-04"))

mean(nums)
mean(dates)
```

Selecting the appropriate method based on the type is called **method dispatch**.

## More on R's class systems

* S3: Informal system where the `class` attribute is set on the object to say what it is. A common convention is to name methods `function.class` (e.g., `mean.Date`) but this is not enforced.
* S4: Formal class system with more work to set up but more guarantees. Method dispatch can be on more than one argument's object type.
* RC/R6: "reference" classes that solve a particular problem related to sharing memory. 

You can do a lot with S3 and upgrade to S4 when you want more formality. Reference classes are rarely a first choice, but can be useful in certain circumstances.

## S3: setting attributes of existing things

The `factor` type in R is a special version of an integer vector:

```{r}
f <- factor(c("a", "b", "c"))

typeof(f)
class(f)
attributes(f)
```

We can remove the `class` attribute and it becomes an integer vector:
```{r}
i <- unclass(f)

typeof(i)
class(i)
attributes(i)
```

Using the `print` generic function on both:
```{r}
print(f)
print(i)
```

If we wanted to fully remove the factor bits:
```{r}
i2 <- i
attr(i2, "levels") <- NULL
print(i2)
```

## Exercise

What class of object does the following code return? What base type is it built on (`typeof`)? What attributes does it use?

```{r exercise-table-class, exercise = TRUE}
x <- table(rpois(100, 5))
x
```

## Making our own S3 classes

As an informal system, we create S3 objects as we need them using the `class<-` function. A really common existing object to build on is a `list`.

```{r}
# Create and assign class in one step
x <- structure(list(), class = "my_class")
class(x)

# Create, then set class
x <- list()
class(x) <- "my_class"
class(x)
```

E.g. Creating a class for students in a course:

```{r}
gordy <- list(name = c(First = "Gordy", Last = "Bluestone"),
          major = "Data Science",
          transcript = list(FA23 = c("DATASCI101" = "A-",
                                     "PHIL101" = "B+"),
                            WN24 = c("MATH113" = "A")))

class(gordy) <- "student"
gordy$name
gordy$transcript
```

## Constructors

We often write **constructor functions** to generate new objects of a given type:

```{r}
make_student <- function(first, last, major, transcript) {
  x <- list(name = c(First = first, Last = last),
          major = major,
          transcript = transcript)
  class(x) <- "student"
  
  return(x)
}

gordy <- make_student("Gordy", "Bluestone", "Data Science",
             list(FA23 = c("DATASCI101" = "A-",
                           "PHIL101" = "B+"),
                  WN24 = c("MATH113" = "A")))

ann <- make_student("Ann", "Otherstudent", "Econ",
             list(WN25 = c("ECON101" = "A")))
```

It might make sense to make transcripts a class as well.

### Writing a method

Let's use the `mean` function to compute the student's GPA.

```{r}
formals(mean)
```


```{r}
honor_points <- c(
  "A+" = 4.0,
  "A" = 4.0,
  "A-" = 3.7,
  "B+" = 3.3,
  "B" = 3.0,
  "B-" = 2.7,
  "C+" = 2.3,
  "C" = 2.0,
  "C-" = 1.7,
  "D+" = 1.3,
  "D" = 1.0,
  "D-" = 0.7,
  "E" = 0.0,
  "P" = 0.0,
  "F" = 0.0,
  "CR" = 0.0,
  "NC" = 0.0
)

# create the function with the same signature as `mean`
mean.student <- function(x, ...) {
  ts <- x$transcript |> unlist()
  points <- honor_points[ts]
  mean(points)
}

mean.student(gordy)
mean.student(ann)
```


```{r}
mean(gordy)
methods(mean)
```

## Writing our own generic functions

Suppose we needed a function to generate a greeting for someone.

```{r}
greet <- function(x) {
  UseMethod("greet")
}

methods(greet)
```

We can set a default:
```{r}
greet.default <- function(x) {
  paste("Hello,", format(x))
}

greet(3)
greet(gordy)
```

## Improving with a method

```{r}
greet.student <- function(x) {
  name <- paste(x$name, collapse = " ")
  greet(name)
}

greet(gordy)
```

## Inheritance: using existing classes

Let's create an new class for alumni:

```{r}
make_alumnus <- function(student, year) {
  student$graduated <- year
  class(student) <- c("alumnus", class(student))
  student
}

grad_gordy <- make_alumnus(gordy, 2025)

print(grad_gordy)
mean(grad_gordy)
greet(grad_gordy)
```

Showing more respect:

```{r}
greet.alumnus <- function(x) {
  v <- NextMethod()
  paste(v, "-- Class of ", x$graduated)
}

greet(gordy)
greet(grad_gordy)
```

## S4 Classes

With S4 classes, we define exactly what fields we want:

```{r}
setClass("s4student",
           slots = c(name = "character",
                     transcript = "list"))

setClass("s4alum", 
           contains = "s4student",
           slots = c(graduated = "integer"))

```

This will provide more checking (e.g., can't give a character as the graduated year), but also more up-front work. Still useful to write constructors and other related functions.

More details on dispatch other topics available in [Advanced R chapter 15](https://adv-r.hadley.nz/s4.html)




