---
subtitle: "Stats 306: Lecture 16"
title: "Strings: Regular Expressions"
output: 
  learnr::tutorial:
    progressive: true
    css: css/lecture.css
runtime: shiny_prerendered
---


```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
set.seed(2939394)
if (!require(nycflights13)) install.packages("nyclfights13", repos = "http://cran.us.r-project.org")

sonnet18 <- "Shall I compare thee to a summer's day?\nThou art more lovely and more temperate:\nRough winds do shake the darling buds of May,\nAnd summer's lease hath all too short a date;\nSometime too hot the eye of heaven shines,\nAnd often is his gold complexion dimm'd;\nAnd every fair from fair sometime declines,\nBy chance or nature's changing course untrimm'd;\nBut thy eternal summer shall not fade,\nNor lose possession of that fair thou ow'st;\nNor shall death brag thou wander'st in his shade,\nWhen in eternal lines to time thou grow'st:\n\tSo long as men can breathe or eyes can see,\n\tSo long lives this, and this gives life to thee."

sonnet18_lines <- str_split(sonnet18, "\n")[[1]]
```

## Review: Strings basics

* Strings vs. factors: strings for text search/combine, factors for non-lexicographic ordering
* Escaping: `\"`, `\\`, `\n` (newline), `\t` (tab)
* Combining: `paste`, `paste0`, `str_c`; use `collapse` for single string from vector
* `str_length`, `str_sub` (retrieve/replace substrings)
* `str_split` returns a list; use `[[1]]` to get vector for a single string

## Looking for patterns: Regexes

A **REG**ular **EX**pression (**regex**) is a mini-language for writing patterns to match in strings.

They are used in R but also in other languages, the command line, databases, etc.

A regex is just a string, albeit one that looks rather strange.

## Regexes pattern matches

Alphanumeric characters (a-z, A-Z, 0-9) match directly, most other characters with some later exceptions.

```{r}
str_view(sonnet18_lines, "thee", html = TRUE)
```

`.` will match exactly one character (not newline)

```{r}
str_view(sonnet18_lines, "th..", html = TRUE)
```

Limited choices with `[]`
```{r}
str_view(sonnet18_lines, "th[aeiou][aeiou]", html = TRUE)
```

Or classes of characters
```{r}
str_view(sonnet18_lines, "[A-Z]", html = TRUE)
```

Negation choices with `[^]`
```{r}
str_view(sonnet18_lines, "th[aeiou][^aeiou ]", html = TRUE)
```

## Other specials

To match periods we need to escape with `\\`

```{r}
str_view(sonnet18_lines[14], "\\.", html = TRUE)
```

Beginning of string is "^", end of string is "$" (can get literals with escapes):

```{r}
str_view(sonnet18_lines, "^.. ", html = TRUE)
```

Any white space (`\\s`) and make it optional `?`
```{r}
str_view(sonnet18_lines, "^\\s?.. ", html = TRUE)
```

`\\d` matches digits.

## Repetition

* `?`: zero or one occurrence
* `*`: zero or many
* `+`: one or more

```{r}
str_view(sonnet18_lines, "m+", html = TRUE)
```

## Groups and Alternation

To select between alternates use a vertical bar `|`. We often combine this with groups based on `()`:

```{r}
str_view(sonnet18_lines, "Thou|thou", html = TRUE)
```

```{r}
str_view_all(sonnet18_lines, "fa(ir|de)", html = TRUE)
```

Repeat patterns in `( )` using `\\NUMBER` (**backreferences**). Contiguous words with same letter: 
```{r}
str_view(sonnet18_lines, "(\\s|^)(.)[^ ]* \\2", html = TRUE)
```
(notice "from fair" does not match because "fair from" already matched)

## Exercise

Find all contractions (words with "'" in them, but not possessive) in `sonnet18_lines` using `str_view`.

```{r contractions, exercise=TRUE}

```

## Using regular expressions

* Apply regexes to column names or levels
* Finding matches in columns
* Extract information from strings
* Break up strings (tokenize) on more complicated patterns

## Art and Emotion data set

```{r}
wikiart <- read_tsv("./data/WikiArt-Emotions-All.tsv.gz")
colnames(wikiart)
```

## Columns

With lots of columns, we could write things out direct, but better to use regex to find columns matching a pattern:

```{r}
select(wikiart, matches("anger"))
select(wikiart, matches("[tT]itle"))
select(wikiart, matches("[tT]itle."))
```

## Making more tidy

This has lots of **ratings**, let's put them into better columns.

```{r}
pivot_longer(wikiart, 
             matches(":"), 
             names_to = c("rating_type", "emotion"),
             names_sep = ":",
             values_to = "mean_annotation") -> wa_long

select(wa_long, Title, rating_type, emotion, mean_annotation)
```

## Find and replace

Hard to see in previous but there are spaces in the emotions:

```{r}
wa_long$emotion[1:5]
```

Let's remove all white space
```{r}
wa_long <- mutate(wa_long,
                  mean_annotation = str_replace(emotion, "\\s", ""))
wa_long$emotion[1:5]
```

## Improving rating type

```{r}
count(wa_long, rating_type)
```

Let's turn that into two columns:
```{r}
wa_long <- mutate(wa_long,
                  presented_image = str_detect(rating_type, "[iI]mage"),
                  presented_title = str_detect(rating_type, "[tT]itle"))

group_by(wa_long, rating_type) |> 
  summarize(sum(presented_image), sum(presented_title))
```





